DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2021-11-02 manuscript/manuscript.Rmd
figures

keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
wordcount         : "2628"

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf

@@

\begin{enumerate}
\item It starts with an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. The tree can be provided in newick format, also as a character string, or as a "phylo" R object, and can have any type of branch lengths or none. 
\item The input taxon names are cleaned with TNRS and saved as a `datelifeQuery` object. If taxon names are taxonomic groups above the species level, `datelife` has two alternative behaviors. If the "get species from taxon" flag is active, `datelife` will retrieve all species within a higher taxon name and add the species names to the input. If the flag is inactive, `datelife` will drop the higher taxon names from the input. The cleaned input taxon names are searched across the source chronogram database. Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms. 
<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!-- This was generating some confusion, so I'm calling them pruned source chronograms now-->
Then, each pruned source chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
\item  Finally, time of lineage divcergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names. %<!--, a taxonomic tree-->
\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
\end{enumerate}

>>>>

keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
wordcount         : "2650"

@@

documentclass     : "apa6"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    includes      :
      # in_header: preamble.tex
      after_body  : [figures.tex]

@@

\begin{enumerate}
\item It starts with an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. The tree can be provided in newick format, also as a character string, or as a "phylo" R object, and can have any type of branch lengths or none. 
\item The input taxon names are cleaned with TNRS and saved as a `datelifeQuery` object. If taxon names are taxonomic groups above the species level, `datelife` has two alternative behaviors. If the "get species from taxon" flag is active, `datelife` will retrieve all species within a higher taxon name and add the species names to the input. If the flag is inactive, `datelife` will drop the higher taxon names from the input. The cleaned input taxon names are searched across the source chronogram database. Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms.
Then, each pruned source chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
\item  Finally, time of lineage divcergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names. %<!--, a taxonomic tree-->
\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
\end{enumerate}

### 2021-11-02 manuscript/manuscript.tex
figures

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 2628}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 2650}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

% in this file I leave the figure captions outside\ caption{} because I want them
to be formatted in the same way as the general text (double spaced and linenumbered)
\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}

\begin{figure}[!h]
\includegraphics{../figures/Fig1.pdf}
\caption{Stylized DateLife workflow. This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{www.datelife.org/query/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
\label{fig:workflow}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:workflow}}
% \end{center}
% Stylized DateLife workflow. This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names. We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray). We also performed a search using the already processed query for comparison (light blue).}
\label{fig:runtime1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:runtime1}}
% \end{center}
%Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names. We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray). We also performed a search using the already processed query for comparison (light blue).
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
\caption{Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).
\label{fig:schronograms1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:schronograms1}}
% \end{center}
%Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).
\begin{figure}[!h]
\includegraphics{../figures/fig_summaries.pdf}
\caption{LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.}
\label{fig:summaries}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:summaries}}
% \end{center}
% LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_bladj.pdf}
\caption{LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.}
\label{fig:cvbladj}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbladj}}
% \end{center}
% LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_boldsumm.pdf}
\caption{LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.}
\label{fig:cvbold}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}
FIGURES
%LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.

%\end{linenumbers}
